Computing sequence likelihoods

### 2017-04-22 writeups/check-likelihood.Rmd
a bit of reordering; un-exported a few things

Consider the case $k=1$, i.e., Exponential.
Let $\lambda = \max_i -G_{ii}$ and $Q_{ij} = G_{ij}/\lambda$ and $Q_{ii}=-\sum_{j \neq i} G_{ij}$,
so that the process jumps as $Q$ at the times of a Poisson($\lambda$) process.
As described in [context/writeups/gamma-prob.html](the context project),
to move from an Exponential($1/\theta$) time to an Exponential($1/(\theta+u)$) time,
we need to add either zero jumps (with probability $\theta/(\theta+u)$)
or a Geometric($\lambda(\theta+u)/(1+\lambda(\theta+u))$) number of jumps.
This is done by `extend.expAtv`;
a similar thing could be done for Gamma times with shape not equal to 1.
```{r infer_expl_time_fun, cache=TRUE}
e_lhoods <- function (opt, config, seq, dscale, ndscale) {
    ans <- numeric(ndscale)
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    # subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=dscale, time="gamma")  # add scale dscale
    subtransmatrix <- projmatrix
    scale <- 0
    for (k in 1:ndscale) {
        subtransmatrix <- extend.expAtv( A=genmatrix, scale=scale, dscale=dscale, gAtv=subtransmatrix )  # add scale dscale
        # return log-likelihood (*not* negative log-likelihood)
        ans[k] <- sum( counts@counts * log(subtransmatrix) )
    }
    return(ans)
}
lower_opt <- function (x) { x$shortwin <- (x$shortwin-1); x }
```

>>>>

Consider the case $k=1$, i.e., Exponential.
Let $\lambda = \max_i -G_{ii}$ and $Q_{ij} = G_{ij}/\lambda$ and $Q_{ii}=-\sum_{j \neq i} G_{ij}$,
so that the process jumps as $Q$ at the times of a Poisson($\lambda$) process.
As described in [context/writeups/gamma-prob.html](the context project),
to move from an Exponential($1/\theta$) time to an Exponential($1/(\theta+u)$) time,
we need to add either zero jumps (with probability $\theta/(\theta+u)$)
or a Geometric($\lambda(\theta+u)/(1+\lambda(\theta+u))$) number of jumps.
This is done by `extend.expAtv`;
a similar thing could be done for Gamma times with shape not equal to 1.
```{r infer_expl_time_fun, cache=TRUE}
e_lhoods <- function (opt, config, seq, dscale, ndscale) {
    ans <- numeric(ndscale)
    longpats <- getpatterns(opt$longwin,config$bases)
    shortpats <- getpatterns(opt$shortwin,config$bases)
    counts <- counttrans.list( list(longpats,shortpats), 
        seqlist=seq[c("initseq","finalseq")],
        leftwin=opt$leftwin, bases=config$bases, 
        shift=0, cyclic=TRUE )
    genmatrix <- do.call( makegenmatrix, c( list( patlen=opt$longwin ), boundary="none", config ) )
    projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), leftwin=leftwin(counts), fpatterns=colnames(counts) )
    # subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=dscale, time="gamma")  # add scale dscale
    subtransmatrix <- projmatrix
    scale <- 0
    for (k in 1:ndscale) {
        subtransmatrix <- contextual:::extend.expAtv( A=genmatrix, scale=scale, dscale=dscale, gAtv=subtransmatrix )  # add scale dscale
        # return log-likelihood (*not* negative log-likelihood)
        ans[k] <- sum( counts@counts * log(subtransmatrix) )
    }
    return(ans)
}
lower_opt <- function (x) { x$shortwin <- (x$shortwin-1); x }
```
